Left-linear rules in adhesive categories

### 2024-02-09 icalp/fusione.tex
pass minimo su 2 e 3

\begin{notation}
  We stipulate here some notational conventions which will be used
  throughout this paper.
  \begin{itemize}
  \item
    Given a category $\X$ we will not distinguish notationally
    between $\X$ and its class of objects: so that ``$X\in \X$'' means
    that $X$ belongs to the class of objects of $\X$.

  \item
    If $1$ is a terminal object in a category $\X$, the unique
    arrow $X\to 1$ from another object $X$ will be denoted by
    $!_X$. Similarly, if $0$ is initial in $\X$ then $?_X$ will denote
    the unique arrow $0\to
    X$. %When $\X$ is $\Set$ and $1$ is a singleton, $\delta_x$ will denote the arrow $1\to X$ with value $x\in X$.

@@

  \item Given an integer $n\in \mathbb{Z}$, $[0,n]$ will denote the
    set natural numbers less or etual than $n$.  Indeed, if $n<0$,
    then $[0,n]=\emptyset$.
  \end{itemize}
\end{notation}

@@

\begin{definition}
  Let $\X$ be a category and $\mathcal{A}$ a subclass of
  $\mor(\X)$. We say that $\mathcal{A}$ is:
  \begin{itemize}
    \parbox{11cm}{\item
      \emph{stable under pushouts (pullbacks)} if for every pushout (pullbacks) square as the one on the right, 	if $m \in \mathcal{A}$ ($n\in \mathcal{A}$) then $n \in \mathcal{A}$ ($m \in \mathcal{A}$);
    \item \emph{closed under composition} if $g, f\in \mathcal{A}$ implies $g\circ f\in \mathcal{A}$ whenever $g$ and $f$ are composable.{\tiny }}\hfill
    \parbox{2cm}{$\xymatrix{A\ar[r]^f  \ar[d]_{m}& B \ar[d]^n \\ C \ar[r]_g & D}$}
    \parbox{11cm}{}\hfill
  \end{itemize}
\end{definition}

@@

\noindent
\parbox{10cm}{
  \begin{definition}[Van Kampen property]
    Let $\X$ be a category and consider the two diagrams aside.

    \hspace{15pt}
    Given  a class of arrows $\mathcal{A}\subseteq \mor(\X)$, we say that the bottom square
    % \emph{has the Van Kampen property relatively to $\mathcal{A}$}, or that it
    is a \emph{$\mathcal{A}$-Van Kampen square} if:
    \begin{enumerate}
    \item is a pushout square;
    \item 	whenever the cube above has pullbacks as back and left faces and the vertical arrows belong to $\mathcal{A}$, then its top face is a pushout if and only if the front and right faces are pullbacks.
    \end{enumerate}
    Pushout squares which enjoy the ``if'' half of item (2) above are called \emph{$\mathcal{A}$-stable}.

@@

\begin{definition}[$\mathcal{M}$-adhesive category]
  Let $\X$ be a category and let $\mathcal{M}$ be a subclass of
  $\mon(\X)$
  % \begin{enumerate}
  % \item $\mathcal{M}$
  including  all isomorphisms, closed under composition,
  % \item $\mathcal{M}$
  and stable under pullbacks and pushouts.
  % \end{enumerate} 
  The category
  % 
  $\X$ is said to be \emph{$\mathcal{M}$-adhesive} if
  \begin{enumerate}
  \item it has \emph{$\mathcal{M}$-pullbacks}, i.e. pullbacks along arrows of $\mathcal{M}$;
  \item it has \emph{$\mathcal{M}$-pushouts}, i.e. pushouts along arrows of $\mathcal{M}$;
  \item  $\mathcal{M}$-pushouts are $\mathcal{M}$-Van Kampen squares.
  \end{enumerate}

@@

\begin{remark}
  \label{rem:salva}
  \emph{Adhesivity} and \emph{quasiadhesivity} as defined in
  \cite{lack2005adhesive,garner2012axioms} coincide with strict
  $\mon(\X) $-adhesivity and strict $\reg(\X)$-adhesivity,
  respectively.
\end{remark}

@@

\todo{P: qui probabilmente si puo' compattare tenendo solo cio' che si usa}
\todo{Si usa tutto per la classe B+ [DC]}
$\mathcal{M}$-adhesivity is well-behaved with respect to  the comma construction \cite{mac2013categories}, as shown by the following theorem.
\begin{theorem}[\cite{ehrig2006fundamentals,lack2005adhesive}]\label{lem:comma}
  Let $\A$ and $\B$ be respectively an $\mathcal{M}$-adhesive and an
  $\mathcal{M}'$-adhesive category. Let also $L:\A\rightarrow \C$ be a
  functor that preserves $\mathcal{M}$-pushouts, and
  $R:\B\rightarrow \C$ be a functor which preserves pullbacks.Then
  $\comma{L}{R}$ is $\cma{M}{M'}$-adhesive, where
  \[
    \cma{M}{M}':=\{(h,k)\in \mathcal{A}(\comma{L}{R}) \mid h\in
    \mathcal{M}, k\in \mathcal{M}'\}\]
\end{theorem}

@@

\begin{corollary}\label{cor:slice}
  Let $X$ be an object of an $\mathcal{M}$-adhesive category
  $\X$. Then $\X/X$ and $X/\X$ are, respectively, is $\mathcal{M}/X$-
  and $X/\mathcal{M}$-adhesive, where
  \[\mathcal{M}/X:=\{m\in \mathcal{A}(\X/X) \mid m\in \mathcal{M} \}
    \qquad X/\mathcal{M}:=\{m\in \mathcal{A}(X/\X) \mid m\in
    \mathcal{M} \}
  \]
\end{corollary}

Another categorical construction which preserves
$\mathcal{M}$-adhesivity property is the formation of the category of
functors.

\begin{theorem}[\cite{ehrig2006fundamentals,lack2005adhesive}]
  \label{thm:functors}
  If $\X$ is an $\mathcal{M}$-adhesive category, then for every small
  category $\Y$, the category $\X^\Y$ of functors $\Y\to \X$ is
  $\mathcal{M}^{\Y}$-adhesive, where
  \[\mathcal{M}^{\Y}:=\{\eta \in \mathcal{A}(\X^\Y) \mid \eta_Y \in
    \mathcal{M} \text{ for every } Y\in \Y\}\]
\end{theorem}

@@

\begin{definition}[\cite{habel2012mathcal,heindel2009category}]
  Let $\X$ be a $\mathcal{M}$-adhesive category, a \emph{left
    $\mathcal{M}$-linear} rule $\rho$ is a pair $(l,r)$ of arrows with
  $l: K \to L$ and $r: K \to R$, such that $l$ belongs to
  $\mathcal{M}$.  The rule $\rho$ is said to be
  \emph{$\mathcal{M}$-linear} if $r\in \mathcal{M}$ too. A rule $\rho$
  is said to be \emph{consuming} if $l$ is not an isomorphism.  The
  object $L$ is the \emph{left-hand side}, $R$ is the \emph{right-hand
    side} and $K$ the \emph{glueing object}.

@@

\full{
  %
  If we look to direct derivations as
  transitions, it is natural to consider them as edges in a direct
  graph. Taking objects as vertices objects led us to the following
  definition \cite{heindel2009category}.

  \begin{definition}
    Let $(\X, \R)$ be a left-linear DPO rewrityng system, with $\X$
    $\mathcal{M}$-adhesive. The \emph{DPO-derivation graph} of
    $(\X, \R)$ is the (large) directed graph $\gpo$ having as vertices
    the objects of $\X$ and in which an edge between $G$ and $H$ is a
    direct derivation $\dder{D}\colon G\Mapsto H$. A \emph{derivation}
    $\der{D}$ between two objects $G$ and $H$ is a path between them
    in $\gpo$. The \emph{source} and \emph{target} of $\der{D}$ are,
    respectively, $G$ and $H$.
  \end{definition}
}

A derivation can now defined simply as a sequence of direct derivations. More precisely:

\begin{definition}[Derivation]
  Let $(\X, \R)$ be a left-linear DPO rewrityng system, with $\X$
  $\mathcal{M}$-adhesive. Given $G, H \in \X$, a \emph{derivation}
  $\der{D}$ with source $G$ and target $H$, written
  $\der{D}: G \Mapsto H$ is defined as a sequence
  $\{\dder{D}_i\}_{i=0}^n$ of direct derivations such that
  \begin{enumerate}
  \item for every index $i$, $\dder{D}_i$ is a direct derivation $G_i \Mapsto G_{i+1}$;
  \item $G_0=G$ and $G_{n+1}=H$.
  \end{enumerate}
  Moreover, for each $G \in \X$ we have $G : G \Mapsto G$ is a
  derivation, called an \emph{empty derivation}.

@@

\noindent
\parbox{8cm}{
  \begin{definition}
    Let $(\X, \R)$ be a left-linear DPO rewriting system, an
    \emph{abstraction equivalence} between two derivations $\der{D}$
    and $\der{D'}$ with the same length and such that
    $r(\der{D})=r(\der{D}')$, is a family of isomorphisms
    $\{\phi_X\}_{X\in \Deltamin(\der{D})}$ such that, for every
    $i\in [0, \lgh(\der{D})]$ the diagram on the right commutes. We
    will say that $\der{D}$are $\der{D}'$ are \emph{abstraction
      equivalent}, if there exists an abstraction equivalence between
    them.
  \end{definition}
}
\parbox{4cm}{\vspace{-1.35em}$\xymatrix@C=40pt{G'_i&D'_i \ar[r]^{g'_i}
    \ar@{>->}[l]_{f'_i}&G'_{i+1}\\ L_i \ar[u]^{m'_i} \ar[d]_{m_i}& K_i
    \ar[u]^{k'_i} \ar[d]_{k_i} \ar[r]^{r_i} \ar@{>->}[l]_{l_i}
    &R\ar[u]^{h'_i} \ar[d]_{h_i}\\G_i
    \ar@/_.45cm/[uu]_(.35){\phi_{G_i}}|\hole&D_i\ar@{>->}[l]^{f_i}\ar@/_.45cm/[uu]_(.35){\phi_{D_i}}|\hole
    \ar[r]_{g_i}&G_{i+1}\ar@/_.45cm/[uu]_{\phi_{H_i}}}$}

@@

\begin{restatable}{proposition}{propUnique}
  \label{prop:unique}
  Let $\X$ be a $\mathcal{M}$-adhesive category. Suppose that the two
  direct derivations $\dder{D}$ and $\dder{D}'$ below, with the same
  match and applying the same left-linear rule $\rho$ are given.
  \[\xymatrix{L \ar[d]_{m}& K \ar[d]^{k}\ar@{>->}[l]_{l} \ar[r]^{r} &
      R \ar[d]^{h} & L \ar[d]_{m}& K \ar[d]^{k'}\ar@{>->}[l]_{l}
      \ar[r]^{r} & R \ar[d]^{h'}\\G & \ar@{>->}[l]^{f} D \ar[r]_{g}& H
      & G & \ar@{>->}[l]^{f'} D' \ar[r]_{g'}& H'}\]
  Then there an abstraction equivalence between $\dder{D}$ and
  $\dder{D}'$, whose first component is $\id{G}$.
\end{restatable}

@@

In this section we discuss the notion of independence between
consecutive rewriting steps, which is a fundamental ingredient of the
rewriting theory, which comes into play when viewing a derivation
sequence as a concurrent computation, the idea being that of
abstracting out the order of execution of independent rewtings. We
observe that the shift to left-linear rules leads to the failure of
various basic properties of the classical notion of independence and
we single out a framework in which these can be re-established.

@@

In the DPO approach to rewriting the canonical notion of independence is that of sequential independence, recalled below.

\begin{definition}[Sequential independence]
  \label{de:sequential-independence}
  Let $(\X, \R)$ be a left-linear DPO rewriting system with $\X$ an
  $\mathcal{M}$-adhesive category. Let also
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ be a derivation of length $2$, with
  $\dder{D}_i$ using rule $\rho_i = (l_i,r_i)$.  An \emph{independence
    pair} between $\dder{D}_0$ and $\dder{D}_1$, is a pair of arrows
  $i_0\colon R_0\to D_1$ and $i_1\colon L_1\to D_0$ such that the
  following diagram commutes.
  \[\xymatrix@C=15pt{L_0 \ar[d]_{m_0}&& K_0
      \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} & R_0
      \ar@{.>}@/^.35cm/[drrr]|(.3)\hole_(.4){i_0}
      \ar[dr]|(.3)\hole_{h_0} && L_1 \ar@{.>}@/_.35cm/[dlll]^(.4){i_1}
      \ar[dl]|(.3)\hole^{m_1}& K_1 \ar[d]^{k_1}\ar@{>->}[l]_{l_1}
      \ar[rr]^{r_1} && R_1 \ar[d]^{h_1}\\G_0 && \ar@{>->}[ll]^{f_0}
      D_0 \ar[rr]_{g_0}&& G_1 && \ar@{>->}[ll]^{f_1} D_1
      \ar[rr]_{g_1}&& G_2}\] We will say that $\dder{D}_0$ and
  $\dder{D}_1$ are
  % \emph{weakly sequentially independent} if an independence pair
  % exists. If such independence pair is unique we will say that
  % $\dder{D}$ and $\dder{D'}$ are
  \emph{sequentially independent}.\todo{Eliminata la nozione di weakly
    indepedendent, giusto?}
\end{definition}

@@

\begin{definition}[Switch]
  \label{def:switch}
  Let $(\X, \R)$ be a left-linear DPO rewriting system with $\X$ an
  $\mathcal{M}$-adhesive category and suppose that
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ is a derivation of length $2$ made
  by two sequentially independent derivations $\dder{D}_0$,
  $\dder{D}_1$, with independence pair $(i_0, i_1)$, as in the diagram
  on the left below. A \emph{switch} of $\der{D}$ along $(i_0,i_1)$ is
  a derivation $\der{E}=\{\dder{E}_i\}_{i=0}^1$, using the same rules
  in reverse order, as in the diagram on the right below, 
  \[
    \def\objectstyle{\scriptstyle}\def\labelstyle{\scriptstyle}
    \xymatrix@C=1.1mm@R=8mm{
      {L_0} \ar[d]_{m_0}
      & & {K_0} \ar@{>->}[ll]_{l_0} \ar[rr]^{r_0} \ar[d]_{k_0}
      & & {R_0} \ar[drr]_{h_0}  \ar@(r,ul)@{.>}^(.75){i_0}[drrrrrr]
      & & & & 
      % 
      {L_1}\ar[dll]^{m_1} \ar@(l,ur)@{.>}_(.75){i_1}[dllllll]
      & & {K_1} \ar@{>->}[ll]_{l_1} \ar[rr]^{r_1} \ar[d]^{k_1}
      & & {R_1} \ar[d]^{h_1} \\
      % 
      % 
      {G}
      & & {D_0} \ar@{>->}[ll]^{f_0} \ar[rrrr]_{g_0}
      & & & & {H} & &
      & &  {D_1} \ar@{>->}[llll]^{f_1} \ar[rr]_{g_1}
      & & {H}
    }
    % 
    \xymatrix@C=1.1mm@R=8mm{
      {L_1} \ar[d]_{m_1'}
      & & {K_1} \ar@{>->}[ll]_{l_1} \ar[rr]^{r_1} \ar[d]_{k_1'}
      & & {R_1} \ar[drr]_{h_1'}  \ar@(r,ul)@{.>}^(.75){i_1'}[drrrrrr]
      & & & & 
      % 
      {L_0}\ar[dll]^{m_0'} \ar@(l,ur)@{.>}_(.75){i_0'}[dllllll]
      & & {K_0} \ar@{>->}[ll]_{l_0} \ar[rr]^{r_0} \ar[d]^{k_0'}
      & & {R_0} \ar[d]^{h_0'} \\
      % 
      % 
      {G}
      & & {D_1'} \ar@{>->}[ll]^{f_1'} \ar[rrrr]_{g_1'}
      & & & & {H'} & &
      & &  {D_0'} \ar@{>->}[llll]^{f_0'} \ar[rr]_{g_0'}
      & & {H'}  }
    %
  \]
  such that there is an independence pair $(i_1', i_0')$ between
  $\dder{E}_0$ and $\dder{E}_1$ and 
  \begin{center}   
    $m_0=f_0' \circ i_0'$
    \qquad $h_1=g_1' \circ i_1'$
    \qquad $m_0'= f_0 \circ i_1$
    \qquad $h_1'= g_{1}\circ i_0$.
  \end{center}

@@

As already mentioned, when viewing derivations as concurrent computation, for linear graph rewriting systems it is canonical to identify  derivations which are equal ``up to switching''. The same notion can be given for left-linear rewriting systems relying on notion of switch.

@@

The fact that sequential indepedence implies switchability always
holds for linear rules. [REF] The result is so fundamental that it is
not even stated, in the sense that switchability is not introduced
axiomatically as in Definition~\ref{def:switch}, but it is based on
the explicit construction of a switch.

When working with left-linear rewriting systems, in general (weakly) sequential independence does not imply switchability (while the converse implication clearly holds), as shown by the example below.

\full{
  \begin{remark}
    \label{rem:fact}
    Notice that, since $f_0'$ is in $\mathcal{M}$, there is
    at most one $i_0' \colon L_0\to D_0'$ such that
    $m_0=f_0' \circ i_0'$. Moreover, we already know that
    there is at most one $i_1' \colon R_1\to D_1'$ such that
    $m_1'=f_1' \circ i_1'$, therefore, the
    independence pair $(i_1',i_0')$ between the component of a switch is
    uniquely determined.
  \end{remark}
}

\begin{example}
  \label{ex:diff1}
  Recall that, given two posets $(P, \leq)$ and $(Q, \leq)$, we can
  construct the poset $(P, \leq)\oplus(Q,\leq)$ which has as
  underlying set the disjoint union $P+Q$ and in which every element
  of $Q$ is greater then every element of $P$, while elements of $P$
  and $Q$ are compared using the original orders.

>>>>

\begin{notation}
  We stipulate some notational conventions that will be used
  throughout the paper.
  \begin{itemize}
  \item
    Given a category $\X$ we will not distinguish notationally
    between $\X$ and its class of objects: so that ``$X\in \X$'' means
    that $X$ belongs to the class of objects of $\X$.

  \item
    If $1$ is a terminal object in $\X$, the unique
    arrow $X\to 1$ from an object $X$ will be denoted by
    $!_X$. Similarly, if $0$ is initial in $\X$ then $?_X$ will denote
    the unique arrow $0\to
    X$. %When $\X$ is $\Set$ and $1$ is a singleton, $\delta_x$ will denote the arrow $1\to X$ with value $x\in X$.

@@

  \item Given an integer $n\in \mathbb{Z}$, $[0,n]$ will denote the
    set of natural numbers less or equal than $n$, assuming that
    $[0,n]=\emptyset$ whenever $n<0$.
  \end{itemize}
\end{notation}

@@

\begin{definition}
  Let $\X$ be a category and $\mathcal{A}$ a subclass of
  $\mor(\X)$. We say that $\mathcal{A}$ is
  \begin{itemize}
    \parbox{11cm}{\item
      \emph{stable under pushouts (pullbacks)} if for every pushout (pullbacks) square as the one on the right, 	if $m \in \mathcal{A}$ ($n\in \mathcal{A}$) then $n \in \mathcal{A}$ ($m \in \mathcal{A}$);
    \item \emph{closed under composition} if $g, f\in \mathcal{A}$ implies $g\circ f\in \mathcal{A}$ whenever $g$ and $f$ are composable.{\tiny }}\hfill
    \parbox{2cm}{$\xymatrix{A\ar[r]^f  \ar[d]_{m}& B \ar[d]^n \\ C \ar[r]_g & D}$}
    \parbox{11cm}{}\hfill
  \end{itemize}
\end{definition}

@@

\noindent
\parbox{10cm}{
  \begin{definition}[Van Kampen property]
    Let $\X$ be a category and consider the two diagrams aside.
 %   
%    \hspace{15pt}
    Given  a class of arrows $\mathcal{A}\subseteq \mor(\X)$, we say that the bottom square
    % \emph{has the Van Kampen property relatively to $\mathcal{A}$}, or that it
    is a \emph{$\mathcal{A}$-Van Kampen square} if
    \begin{enumerate}
    \item is a pushout square;
    \item 	whenever the cube above has pullbacks as back and left faces and the vertical arrows belong to $\mathcal{A}$, then its top face is a pushout 
    if and only if the front and right faces are pullbacks.
    \end{enumerate}
    Pushout squares which enjoy the ``if'' half of item (2) above are called \emph{$\mathcal{A}$-stable}.

@@

\begin{definition}[$\mathcal{M}$-adhesive category]
  Let $\X$ be a category and $\mathcal{M}$ a subclass of
  $\mon(\X)$
  % \begin{enumerate}
  % \item $\mathcal{M}$
  including  all isomorphisms, closed under composition,
  % \item $\mathcal{M}$
  and stable under pullbacks and pushouts.
  % \end{enumerate} 
  The category
  % 
  $\X$ is said to be \emph{$\mathcal{M}$-adhesive} if
  \begin{enumerate}
  \item it has \emph{$\mathcal{M}$-pullbacks}, i.e. pullbacks along arrows of $\mathcal{M}$;
  \item it has \emph{$\mathcal{M}$-pushouts}, i.e. pushouts along arrows of $\mathcal{M}$;
  \item  $\mathcal{M}$-pushouts are $\mathcal{M}$-Van Kampen squares.
  \end{enumerate}

@@

\begin{remark}
  \label{rem:salva}
  \emph{Adhesivity} and \emph{quasiadhesivity} 
  \cite{lack2005adhesive,garner2012axioms} coincide with strict
  $\mon(\X) $-adhesivity and strict $\reg(\X)$-adhesivity,
  respectively, for $\reg(\X)$ the class of regular monos.
\end{remark}

@@

\todo{P: qui probabilmente si puo' compattare tenendo solo cio' che si usa}
\todo{Si usa tutto per la classe B+ [DC]}
$\mathcal{M}$-adhesivity is well-behaved with respect to  the comma construction \cite{mac2013categories}, as shown by the following theorem.
\begin{theorem}
[\cite{ehrig2006fundamentals,lack2005adhesive}]
\label{lem:comma}
  Let $\A$ and $\B$ be respectively an $\mathcal{M}$-adhesive and an
  $\mathcal{M}'$-adhesive category. Let also $L:\A\rightarrow \C$ be a
  functor that preserves $\mathcal{M}$-pushouts, and
  $R:\B\rightarrow \C$ a functor that preserves pullbacks.Then
  $\comma{L}{R}$ is $\cma{M}{M'}$-adhesive, where
  \[
    \cma{M}{M}':=\{(h,k)\in \mathcal{A}(\comma{L}{R}) \mid h\in
    \mathcal{M}, k\in \mathcal{M}'\}\]
\end{theorem}

@@

\begin{corollary}\label{cor:slice}
  Let $X$ be an object of an $\mathcal{M}$-adhesive category
  $\X$. Then $\X/X$ and $X/\X$ are, respectively, $\mathcal{M}/X$-
  and $X/\mathcal{M}$-adhesive, where
  \[\mathcal{M}/X:=\{m\in \mathcal{A}(\X/X) \mid m\in \mathcal{M} \}
    \qquad X/\mathcal{M}:=\{m\in \mathcal{A}(X/\X) \mid m\in
    \mathcal{M} \}
  \]
\end{corollary}

Another construction that preserves
$\mathcal{M}$-adhesivity is the formation of the functor category.

\begin{theorem}[\cite{ehrig2006fundamentals,lack2005adhesive}]
  \label{thm:functors}
  Let $\X$ be an $\mathcal{M}$-adhesive category. Then for every small
  category $\Y$, the category $\X^\Y$ of functors $\Y\to \X$ is
  $\mathcal{M}^{\Y}$-adhesive, where
  \[\mathcal{M}^{\Y}:=\{\eta \in \mathcal{A}(\X^\Y) \mid \eta_Y \in
    \mathcal{M} \text{ for every } Y\in \Y\}\]
\end{theorem}

@@

\begin{definition}[\cite{habel2012mathcal,heindel2009category}]
  Let $\X$ be an $\mathcal{M}$-adhesive category. A \emph{left
    $\mathcal{M}$-linear} rule $\rho$ is a pair $(l,r)$ of arrows with
  $l: K \to L$ and $r: K \to R$, such that $l$ belongs to
  $\mathcal{M}$.  The rule $\rho$ is said
  \emph{$\mathcal{M}$-linear} if $r\in \mathcal{M}$ too. A rule $\rho$
  is said to be \emph{consuming} if $l$ is not an isomorphism.  The
  object $L$ is the \emph{left-hand side}, $R$ is the \emph{right-hand
    side}, and $K$ the \emph{glueing object}.

@@

\full{
  %
  If we look to direct derivations as
  transitions, it is natural to consider them as edges in a direct
  graph. Taking objects as vertices leds us to the following
  definition \cite{heindel2009category}.

  \begin{definition}
    Let  $\X$ be an $\mathcal{M}$-adhesive category and 
    $(\X, \R)$ a left-linear DPO rewriting system. 
    The \emph{DPO-derivation graph} of
    $(\X, \R)$ is the (large) directed graph $\gpo$ having as vertices
    the objects of $\X$ and in which an edge between $G$ and $H$ is a
    direct derivation $\dder{D}\colon G\Mapsto H$. A \emph{derivation}
    $\der{D}$ between two objects $G$ and $H$ is a path between them
    in $\gpo$. The \emph{source} and \emph{target} of $\der{D}$ are,
    respectively, $G$ and $H$.
  \end{definition}
}

A derivation can now defined simply as a sequence of direct derivations.

\begin{definition}[Derivation]
  Let  $\X$ be an $\mathcal{M}$-adhesive category and 
  $(\X, \R)$ a left-linear DPO rewriting system. 
  Given $G, H \in \X$, a \emph{derivation}
  $\der{D}$ with source $G$ and target $H$, written
  $\der{D}: G \Mapsto H$ is defined as a sequence
  $\{\dder{D}_i\}_{i=0}^n$ of direct derivations such that
  \begin{enumerate}
  \item for every index $i$, $\dder{D}_i$ is a direct derivation $G_i \Mapsto G_{i+1}$;
  \item $G_0=G$ and $G_{n+1}=H$.
  \end{enumerate}
  Moreover, for each $G \in \X$ we have $G : G \Mapsto G$ is a
  derivation, called an \emph{empty derivation}.

@@

\noindent
\parbox{8cm}{
  \begin{definition}
    Let  $\X$ be an $\mathcal{M}$-adhesive category and 
    $(\X, \R)$ a left-linear DPO rewriting system. 
    An \emph{abstraction equivalence} between two derivations $\der{D}$
    and $\der{D'}$ with the same length and
    $r(\der{D})=r(\der{D}')$, is a family of isomorphisms
    $\{\phi_X\}_{X\in \Deltamin(\der{D})}$ such that the diagram on the right commutes
    for every $i\in [0, \lgh(\der{D})]$. We
    will say that $\der{D}$are $\der{D}'$ are \emph{abstraction
      equivalent} if there exists an abstraction equivalence between
    them.
  \end{definition}
}
\parbox{4cm}{\vspace{-1.35em}$\xymatrix@C=40pt{G'_i&D'_i \ar[r]^{g'_i}
    \ar@{>->}[l]_{f'_i}&G'_{i+1}\\ L_i \ar[u]^{m'_i} \ar[d]_{m_i}& K_i
    \ar[u]^{k'_i} \ar[d]_{k_i} \ar[r]^{r_i} \ar@{>->}[l]_{l_i}
    &R\ar[u]^{h'_i} \ar[d]_{h_i}\\G_i
    \ar@/_.45cm/[uu]_(.35){\phi_{G_i}}|\hole&D_i\ar@{>->}[l]^{f_i}\ar@/_.45cm/[uu]_(.35){\phi_{D_i}}|\hole
    \ar[r]_{g_i}&G_{i+1}\ar@/_.45cm/[uu]_{\phi_{H_i}}}$}

@@

\begin{restatable}{proposition}{propUnique}
  \label{prop:unique}
  Let $\X$ be a $\mathcal{M}$-adhesive category. Suppose that the two
  direct derivations $\dder{D}$ and $\dder{D}'$ below, with the same
  match and applying the same left-linear rule $\rho$ are given
  \[\xymatrix{L \ar[d]_{m}& K \ar[d]^{k}\ar@{>->}[l]_{l} \ar[r]^{r} &
      R \ar[d]^{h} & L \ar[d]_{m}& K \ar[d]^{k'}\ar@{>->}[l]_{l}
      \ar[r]^{r} & R \ar[d]^{h'}\\G & \ar@{>->}[l]^{f} D \ar[r]_{g}& H
      & G & \ar@{>->}[l]^{f'} D' \ar[r]_{g'}& H'}\]
  Then there an abstraction equivalence between $\dder{D}$ and
  $\dder{D}'$, whose first component is $\id{G}$.
\end{restatable}

@@

In this section we discuss the notion of independence between
consecutive rewriting steps, which is a fundamental ingredient of the
rewriting theory, which comes into play when viewing a derivation
sequence as a concurrent computation, the idea being that of
abstracting out the order of execution of independent rewritings. We
observe that the shift to left-linear rules leads to the failure of
various basic properties of the classical notion of independence and
we single out a framework in which these can be re-established.

@@

In the DPO approach the canonical notion of independence is that of sequential independence.
%recalled below.

\begin{definition}[Sequential independence]
  \label{de:sequential-independence}
  Let with $\X$ an $\mathcal{M}$-adhesive category 
  and $(\X, \R)$ a left-linear DPO rewriting system. 
  Let also
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ be a derivation of length $2$, with
  $\dder{D}_i$ using rule $\rho_i = (l_i,r_i)$.  An \emph{independence
    pair} between $\dder{D}_0$ and $\dder{D}_1$, is a pair of arrows
  $i_0\colon R_0\to D_1$ and $i_1\colon L_1\to D_0$ such that the
  following diagram commutes
  \[\xymatrix@C=15pt{L_0 \ar[d]_{m_0}&& K_0
      \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} & R_0
      \ar@{.>}@/^.35cm/[drrr]|(.3)\hole_(.4){i_0}
      \ar[dr]|(.3)\hole_{h_0} && L_1 \ar@{.>}@/_.35cm/[dlll]^(.4){i_1}
      \ar[dl]|(.3)\hole^{m_1}& K_1 \ar[d]^{k_1}\ar@{>->}[l]_{l_1}
      \ar[rr]^{r_1} && R_1 \ar[d]^{h_1}\\G_0 && \ar@{>->}[ll]^{f_0}
      D_0 \ar[rr]_{g_0}&& G_1 && \ar@{>->}[ll]^{f_1} D_1
      \ar[rr]_{g_1}&& G_2}\] We will say that $\dder{D}_0$ and
  $\dder{D}_1$ are
  % \emph{weakly sequentially independent} if an independence pair
  % exists. If such independence pair is unique we will say that
  % $\dder{D}$ and $\dder{D'}$ are
  \emph{sequentially independent}.\todo{Eliminata la nozione di weakly
    indepedendent, giusto?}
\end{definition}

@@

\begin{definition}[Switch]
  \label{def:switch}
  Let with $\X$ an $\mathcal{M}$-adhesive category 
  and $(\X, \R)$ a left-linear DPO rewriting system. 
  Let also
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ be a derivation of length $2$ made
  by two sequentially independent derivations $\dder{D}_0$,
  $\dder{D}_1$, with independence pair $(i_0, i_1)$, as in the diagram
  on the left below. A \emph{switch} of $\der{D}$ along $(i_0,i_1)$ is
  a derivation $\der{E}=\{\dder{E}_i\}_{i=0}^1$, using the same rules
  in reverse order, as in the diagram on the right below, 
  \[
    \def\objectstyle{\scriptstyle}\def\labelstyle{\scriptstyle}
    \xymatrix@C=1.1mm@R=8mm{
      {L_0} \ar[d]_{m_0}
      & & {K_0} \ar@{>->}[ll]_{l_0} \ar[rr]^{r_0} \ar[d]_{k_0}
      & & {R_0} \ar[drr]_{h_0}  \ar@(r,ul)@{.>}^(.75){i_0}[drrrrrr]
      & & & & 
      % 
      {L_1}\ar[dll]^{m_1} \ar@(l,ur)@{.>}_(.75){i_1}[dllllll]
      & & {K_1} \ar@{>->}[ll]_{l_1} \ar[rr]^{r_1} \ar[d]^{k_1}
      & & {R_1} \ar[d]^{h_1} \\
      % 
      % 
      {G}
      & & {D_0} \ar@{>->}[ll]^{f_0} \ar[rrrr]_{g_0}
      & & & & {H} & &
      & &  {D_1} \ar@{>->}[llll]^{f_1} \ar[rr]_{g_1}
      & & {H}
    }
    % 
    \xymatrix@C=1.1mm@R=8mm{
      {L_1} \ar[d]_{m_1'}
      & & {K_1} \ar@{>->}[ll]_{l_1} \ar[rr]^{r_1} \ar[d]_{k_1'}
      & & {R_1} \ar[drr]_{h_1'}  \ar@(r,ul)@{.>}^(.75){i_1'}[drrrrrr]
      & & & & 
      % 
      {L_0}\ar[dll]^{m_0'} \ar@(l,ur)@{.>}_(.75){i_0'}[dllllll]
      & & {K_0} \ar@{>->}[ll]_{l_0} \ar[rr]^{r_0} \ar[d]^{k_0'}
      & & {R_0} \ar[d]^{h_0'} \\
      % 
      % 
      {G}
      & & {D_1'} \ar@{>->}[ll]^{f_1'} \ar[rrrr]_{g_1'}
      & & & & {H'} & &
      & &  {D_0'} \ar@{>->}[llll]^{f_0'} \ar[rr]_{g_0'}
      & & {H'}  }
    %
  \]
  such that there is an independence pair $(i_1', i_0')$ between
  $\dder{E}_0$ and $\dder{E}_1$ and 
  \begin{center}   
    $m_0=f_0' \circ i_0'$
    \qquad $h_1=g_1' \circ i_1'$
    \qquad $m_0'= f_0 \circ i_1$
    \qquad $h_1'= g_{1}\circ i_0$.
  \end{center}

@@

As already mentioned, when viewing derivations as concurrent computations, 
for linear graph rewriting systems it is canonical to identify  derivations which are equal 
``up to switching''. The same notion can be given for left-linear rewriting systems relying on notion of switch.

@@

The fact that sequential indepedence implies switchability always
holds for linear rules. [REF] The result is so fundamental that it is
not even stated explicitly, in the sense that switchability is not introduced
axiomatically as in Definition~\ref{def:switch}, but it is based on
the explicit construction of a switch.

When working with left-linear rewriting systems, sequential independence does not imply 
switchability (while the converse implication clearly holds), as shown by the example below.

\full{
  \begin{remark}
    \label{rem:fact}
    Notice that, since $f_0'$ is in $\mathcal{M}$, there is
    at most one $i_0' \colon L_0\to D_0'$ such that
    $m_0=f_0' \circ i_0'$. Moreover, we already know that
    there is at most one $i_1' \colon R_1\to D_1'$ such that
    $m_1'=f_1' \circ i_1'$, therefore, the
    independence pair $(i_1',i_0')$ between the components of a switch is
    uniquely determined.
  \end{remark}
}

\begin{example}
  \label{ex:diff1}
  Recall that, given two posets $(P, \leq)$ and $(Q, \leq)$, we can
  construct the poset $(P, \leq)\oplus(Q,\leq)$ which has as
  underlying set the disjoint union $P+Q$ and in which every element
  of $Q$ is greater than every element of $P$, while elements of $P$
  and $Q$ are compared using the original orders.
